Shapiro delay

### 2021-03-12 tpe_edreview2_rebuttal.tex

and thus their difference is

\[
\ell_\mathrm{Sch} - \ell_\mathrm{flat}\sim M\log w
\]

which goes to infinity logarithmically; in other words, for an
infinite string the Schwarzschild metric is a ``worse infinity" which
diverges faster than what occurs in flat spacetime.  So far, so good.
I guess what bothers me is the statement that ``even far from [emphasis
  added] the event horizon, radial strings are arbitrarily longer than
flat space would imply...".  So long as $w$ is large (however far from
the event horizon) but finite, the lengths in both metrics are precise
numbers, not arbitrary.  Only when $w$ goes to infinity does the
length in Schwarzschild spacetime become infinitely longer than the
length in flat spacetime, although though both are infinite.  I
suppose this passage would sit better with this reader by merely
stating that the asymptotic difference between $\ell_\mathrm{Sch}$ and
$\ell_\mathrm{flat}$ diverges logarithmically.  But perhaps this is
only a matter of taste.

\textcolor{blue}{This is a subtle point which took me some
  considerable time to understand.  I would express the problem with
  the existing text as that the radial strings concerned stretch all
  the way to the event horizon; yet the phrase ``far from the event
  horizon'' kind of suggests that {\em all} the string is far from the
  event horizon.\\ It is possible to get round this informally by
  considering the proper length of a radial string from $2M(1+w_1)$ to
  $2M(1+w_2)$ (where we understand $2M\leq w_1\leq w_2$), which is seen from
  equation 4 to be $2M\left[w_1-w_2 + \log\left(\frac{w_2}{w_1}\right) +
  O(w_1^{-1})\right]$.\\ But on balance this is a rather unspectacular result
  which would require a distractingly large amount of explanation.  On
  balance I think it best to leave this out, resulting, I hope in
  sharper, more focussed wording.
\\ {\bf old text}
The discrepancy\ldots has the startling feature that
$\lim_{w\longrightarrow\infty}\Delta=\infty$ (because of the $\log w$
term).  This suggests that, even far from the event horizon, radial
strings are arbitrarily longer than flat space would imply: a nice
illustration of the failure of Euclidean geometry.
\\ {\bf new text}
The discrepancy\ldots has the startling feature that
$\lim_{w\longrightarrow\infty}\Delta=\infty$ (because of the $\log w$
term): a nice illustration of the failure of Euclidean geometry.
}

@@

\[
=
\sqrt{r_2(r_2-2M)}-\sqrt{r_1(r_1-2M)} + 2M\log\left|
\frac{\sqrt{r_2}+\sqrt{r_2-2M}}{\sqrt{r_1}+\sqrt{r_1-2M}}\right|
\]

In contrast, Eq (2) reads (points of disagreement in red):

\[
\sqrt{r_2(r_2-2M)}-\sqrt{r_1(r_1-2M)} 
+\color{red}M
\color{black}\log\left|
\frac{\sqrt{r_2}+\sqrt{r_2-2M}}{\sqrt{r_1}
  \color{red}-
  \color{black}\sqrt{r_1-2M}}\right|
\]

{\color{blue} The reviewer's expression above agrees with my own
  calculations, but I would observe that the expression above (with or
  without the red alternative) is {\em not} equation 2 of the
  manuscript.  The reviewer's expression actually matches the RHS of
  Equation 2 in the manuscript which is

### 2021-03-12 tpe_edreview2_rebuttal.tex

{\color{blue} The reviewer's expression above is correct}

In contrast, Eq (2) reads (points of disagreement in red):

\[
\sqrt{r_2(r_2-2M)}-\sqrt{r_1(r_1-2M)} 
+\color{red}M
\color{black}\log\left|
\frac{\sqrt{r_2}+\sqrt{r_2-2M}}{\sqrt{r_1}
  \color{red}-
  \color{black}\sqrt{r_1-2M}}\right|
\]

{\color{blue} The expression above (with or without the red
  alternative) is {\em not} equation 2 of the manuscript. Equation 2
  in the manuscript is:

@@

  \[
  \left.
  \sqrt{r(r-2M)} +M\log\left(
  \frac{\sqrt{r}+\sqrt{r-2M}}{\sqrt{r}-\sqrt{r-2M}}\right)
  \right|_{r_1}^{r_2}
  \]

  \[
  = 
  \sqrt{r_2(r_2-2M)}-  \sqrt{r_1(r_1-2M)}
  +M\log\left(
  \frac{\sqrt{r_2}+\sqrt{r_2-2M}}{\sqrt{r_2}-\sqrt{r_2-2M}}
  \frac{\sqrt{r_1}-\sqrt{r_1-2M}}{\sqrt{r_1}+\sqrt{r_1-2M}}
  \right)
  \]

  \[
  = 
  \sqrt{r_2(r_2-2M)}-  \sqrt{r_1(r_1-2M)}
  +M\log\left(
  \frac{
    \left(\sqrt{r_2}+\sqrt{r_2-2M}\right)^2
  }{
    \left(\sqrt{r_1}+\sqrt{r_1-2M}\right)^2
  }
  \right)
  \]

  \[
  = 
  \sqrt{r_2(r_2-2M)}-  \sqrt{r_1(r_1-2M)}
  +2M\log\left(
  \frac{\sqrt{r_2}+\sqrt{r_2-2M}}{\sqrt{r_1}+\sqrt{r_1-2M}}
  \right)
  \]

We see that my original expression matches that of the reviewer.  The
apparent factor of 2 resolves itself because multiplying numerator and
denominator by the appropriate conjugate surd results in the {\em
  square} of the expression given by the referee.

@@

Arguably the referee's formulation is neater and more compact than
that appearing in the manuscript but I would support my own version on
the grounds that it is closer to mathematica's output (and indeed is
much more amenable to asymptotic analysis).

{\bf Summary: both my, and the reviewer's, algebra was correct and our
  expressions in fact agree.}  }

### 2021-03-14 tpe_edreview2_rebuttal.tex

\documentclass[12pt]{article}
\usepackage{xcolor}
\begin{document}

\section*{Rebuttal to second editorial review: overall response}

(Below, the reviewer's comments are in black, and my replies to the
issues are in \textcolor{blue}{blue} [the reviewer used
  \textcolor{red}{red} to highlight differences between our
  calculations].  Short story: I have accommodated the review's minor
comments with slight rewording; and resolved the algebraic discrepancy
by proving that our two expressions are in fact equal).

\section*{Detailed rebuttal}

### 2021-04-19 shapiro.Rmd
fixes #21

---
title: "Shapiro delay"
output: html_document
---

@@

The Shapiro delay is one of the classical tests of general relativity.
Here I attempt to reproduce that analysis using numerical reasoning
but fail because of numerical roundoff error.

@@

First some setups:
```{r}

source("usefulnullgeodesicfuncs.R")

r0 <- r_sun/d # grazing incidence of the sun in Schwarzschild radii
r0
r_s
```

@@

```{r}
b_radial <- r0/sqrt(1-1/r0)

small <- 1e-6  # needed to prevent numerical problems

@@

(function `nullgeodesic()` is a general-purpose solver for the path
light takes in the Schwarzschild metric; see file
`light_start_at_r_equals_2.R` for a more visually appealing simulation
than presented here, and `light_start_at_radius_of_sun.R` for a
discussion of the bending of starlight by the sun during an eclipse,
as (e.g.) measured by Eddington in 1919).  Above, see how the path
starts at $t=0$ at a location given by the radius of the sun (the `r`
column) and angle $\phi=0$ (the `phi` column).  We can perform the
same calculation but for Venus"

@@

The next step is to approximate the funtion

@@

So now `f_earth(t)` gives the time required to achieve a particular
radius.  Thus to calculate the transit time from Earth to Venus we
simply add the times to get to the Earth's radius to Venus's:

@@

## Now calculate the classical time, simply distance/speed:
transit_time_classical <- (orbital_radius_earth + orbital_radius_venus)/sol # seconds

## plot (rather dull):
st <- function(x){(x-min(x))/(max(x)-min(x))}
plot(out_earth$x,out_earth$y,type='o',col='red',cex=st(out_earth$time),asp=1)
event_horizon()

shapiro_delay <- transit_time_relativistic - transit_time_classical
# compare observation of 200 microseconds (!)

shapiro_delay <- transit_time_relativistic - transit_time_classical
cat("shapiro delay = ", shapiro_delay, "seconds")
```

>>>>

---
title: "Shapiro delay"
output:
  pdf_document: default
  html_document: default
---

@@

The Shapiro delay is one of the classical tests of general relativity.
Here I use direct numerical simulation to

@@

First some setups:
```{r}
source("usefulnullgeodesicfuncs.R")
```

Schwarzschild radius of sun:

```{r}
r_s
```

```{r}
r0 <- r_sun/r_s # grazing incidence of the sun in Schwarzschild radii
r0
```

@@

b_radial <- r0/sqrt(1-1/r0) 
r0
b_radial
small <- 1e-10 # needed to prevent numerical problems

@@

(function `nullgeodesic()` is a general-purpose solver for the path
light takes in the Schwarzschild metric; see file
`light_start_at_r_equals_2.R` for a more visually appealing simulation
than presented here, and `light_start_at_radius_of_sun.R` for a
discussion of the bending of starlight by the sun during an eclipse,
as (e.g.) measured by Eddington in 1919).  Above, see how the path
starts at $t=0$ at a location given by the radius of the sun (the `r`
column) and angle $\phi=0$ (the `phi` column).  We can perform the
same calculation but for Venus:

@@

The next step is to approximate the function

@@

```{r}
st <- function(x){(x-min(x))/(max(x)-min(x))}
plot(out_earth$x,out_earth$y,type='o',col='red',cex=st(out_earth$time),asp=1)
event_horizon()
```

So now `f_earth(t)` gives the time required to achieve a particular
radius.  Thus to calculate the transit time from Earth to Venus (NB:
one way!) we simply add the times to get to the Earth's radius to
Venus's:

@@

Now we calculate the classical time.  The naive estimate would be

```{r}
transit_time_classical <- (orbital_radius_earth + orbital_radius_venus)/sol # seconds
```

but this neglects the fact that the light path starts at grazing
incidence (note that the relativistic path does not graze the sun but
actually grazes radius $r_s+2MG/c^2$, but this is the same to first
order).

![Diagram of light path from Earth to Venus via grazing incidence at sun](earth_venus_sun_eddington_shapiro.png){width=100px}

```{r}
distance_grazing_to_earth <- sqrt(orbital_radius_earth^2 - r_sun^2)
distance_grazing_to_venus <- sqrt(orbital_radius_venus^2 - r_sun^2)
transit_time_classical <- (distance_grazing_to_earth + distance_grazing_to_venus)/sol
transit_time_classical
```

Then the Shapiro delay is just the difference:

```{r}
shapiro_delay <- (transit_time_classical - transit_time_relativistic)*2
shapiro_delay
```

(we multiply by two because the radar had an outward and return leg,
to Venus and back).  Compare this with the measured effect of 200
microseconds:

```{r}
shapiro_delay/200e-6
```
